Codebook
========
Codebook was created on `r as.character(Sys.time())` during the same process that generated the dataset. See `run_analysis.md` or `run_analysis.html` for details on data set creation.

Variable list and descriptions
------------------------------
        
Variable name    | Description
-----------------|------------
subject          | Subject who performed the activity
activity         | Activity name
domain           | Time signal or frequency domain signal (Time or Freq)
instrument       | Measuring instrument (Accelerometer or Gyroscope)
acceleration     | Acceleration signal (Body or Gravity)
variable         | Variable (Mean or SD)
jerk             | Jerk signal
magnitude        | Magnitude of the signals
axis             | 3-axial signals in the X, Y and Z directions (X, Y, or Z)
count            | Count of data points used to compute `average`
average          | Average of each variable for each activity and each subject

Data set structure
------------------
```{r}
str(dtTidy)
```

Key variables in the data set
-----------------------------
        
```{r}
key(dtTidy)
```

Sample of data set rows
-----------------------
        
```{r}
dtTidy
```

Summary of variables
--------------------
        
```{r}
summary(dtTidy)
```

Save to file
------------
        
Save data set objects to a tab-delimited text file called `tidydata.txt`.

```{r save}
f <- file.path(dataDirectory, "tidydata.txt")
write.table(dtTidy, f, quote=FALSE, sep="\t", row.names=FALSE)
```